---
title: middleware.Rmd
output:
  html_document:
    toc: yes
---

This file contains a general toolkit for routing and combining bits of HTTP-handling logic. It is similar in spirit to Rook (and Rack, and WSGI, and Connect, and...) but adds cascading and routing.

This file is called "middleware" because that's the term used for these bits of logic in these other frameworks. However, our code uses the word "handler" so we'll stick to that for the rest of this document; just know that they're basically the same concept.

## Intro to handlers

A **handler** (or sometimes, **httpHandler**) is a function that takes a `req` parameter--a request object as described in the Rook specification--and returns `NULL`, or an `httpResponse`.

```{r}
httpResponse <- function(status = 200,
                         content_type = "text/html; charset=UTF-8",
                         content = "",
                         headers = list()) {
  # Make sure it's a list, not a vector
  headers <- as.list(headers)
  if (is.null(headers$`X-UA-Compatible`))
    headers$`X-UA-Compatible` <- "chrome=1"
  resp <- list(status = status, content_type = content_type, content = content,
               headers = headers)
  class(resp) <- 'httpResponse'
  return(resp)
}
```

You can think of a web application as being simply an aggregation of these functions, each of which performs one kind of duty. Each handler in turn gets a look at the request and can decide whether it knows how to handle it. If so, it returns an `httpResponse` and processing terminates; if not, it returns `NULL` and the next handler gets to execute. If the final handler returns `NULL`, a 404 response should be returned.

We have a similar construct for websockets: **websocket handlers** or **wsHandlers**. These take a single `ws` argument which is the websocket connection that was just opened, and they can either return `TRUE` if they are handling the connection, and `NULL` to pass responsibility on to the next wsHandler.

### Combining handlers

Since it's so common for httpHandlers to be invoked in this "cascading" fashion, we'll introduce a function that takes zero or more handlers and returns a single handler. And while we're at it, making a directory of static content available is such a common thing to do, we'll allow strings representing paths to be used instead of handlers; any such strings we encounter will be converted into `staticHandler` objects.

```{r eval=FALSE}
joinHandlers <- function(handlers) {
  # Zero handlers; return a null handler
  if (length(handlers) == 0)
    return(function(req) NULL)

  # Just one handler (function)? Return it.
  if (is.function(handlers))
    return(handlers)

  handlers <- lapply(handlers, function(h) {
    if (is.character(h))
      return(staticHandler(h))
    else
      return(h)
  })

  # Filter out NULL
  handlers <- handlers[!sapply(handlers, is.null)]

  if (length(handlers) == 0)
    return(function(req) NULL)
  if (length(handlers) == 1)
    return(handlers[[1]])

  function(req) {
    for (handler in handlers) {
      response <- handler(req)
      if (!is.null(response))
        return(response)
    }
    return(NULL)
  }
}
```

Note that we don't have an equivalent of `joinHandlers` for wsHandlers. It's easy to imagine it, we just haven't needed one.

### Handler routing

Handlers do not have a built-in notion of routing. Conceptually, given a list of handlers, all the handlers are peers and they all get to see every request (well, up until the point that a handler returns a response).

You could implement routing in each handler by checking the request's `PATH_INFO` field, but since it's such a common need, let's make it simple by introducing a `routeHandler` function. This is a handler [decorator](http://en.wikipedia.org/wiki/Decorator_pattern) and it's responsible for 1) filtering out requests that don't match the given route, and 2) temporarily modifying the request object to take the matched part of the route off of the `PATH_INFO` (and add it to the end of `SCRIPT_NAME`). This way, the handler doesn't need to figure out about what part of its URL path has already been matched via routing.

(BTW, it's safe for `routeHandler` calls to nest.)

```{r eval=FALSE}
routeHandler <- function(prefix, handler) {
  force(prefix)
  force(handler)

  if (identical("", prefix))
    return(handler)

  if (length(prefix) != 1 || !isTRUE(grepl("^/[^\\]+$", prefix))) {
    stop("Invalid URL prefix \"", prefix, "\"")
  }

  pathPattern <- paste("^\\Q", prefix, "\\E/", sep = "")
  function(req) {
    if (isTRUE(grepl(pathPattern, req$PATH_INFO))) {
      origScript <- req$SCRIPT_NAME
      origPath <- req$PATH_INFO
      on.exit({
        req$SCRIPT_NAME <- origScript
        req$PATH_INFO <- origPath
      }, add = TRUE)
      pathInfo <- substr(req$PATH_INFO, nchar(prefix)+1, nchar(req$PATH_INFO))
      req$SCRIPT_NAME <- paste(req$SCRIPT_NAME, prefix, sep = "")
      req$PATH_INFO <- pathInfo
      return(handler(req))
    } else {
      return(NULL)
    }
  }
}
```

We have a version for websocket handlers as well. Pity about the copy/paste job.

```{r eval=FALSE}
routeWSHandler <- function(prefix, wshandler) {
  force(prefix)
  force(wshandler)

  if (identical("", prefix))
    return(wshandler)

  if (length(prefix) != 1 || !isTRUE(grepl("^/[^\\]+$", prefix))) {
    stop("Invalid URL prefix \"", prefix, "\"")
  }

  pathPattern <- paste("^\\Q", prefix, "\\E/", sep = "")
  function(ws) {
    req <- ws$request
    if (isTRUE(grepl(pathPattern, req$PATH_INFO))) {
      origScript <- req$SCRIPT_NAME
      origPath <- req$PATH_INFO
      on.exit({
        req$SCRIPT_NAME <- origScript
        req$PATH_INFO <- origPath
      }, add = TRUE)
      pathInfo <- substr(req$PATH_INFO, nchar(prefix)+1, nchar(req$PATH_INFO))
      req$SCRIPT_NAME <- paste(req$SCRIPT_NAME, prefix, sep = "")
      req$PATH_INFO <- pathInfo
      return(wshandler(ws))
    } else {
      return(NULL)
    }
  }
}
```

### Handler implementations

Now let's actually write some handlers. Note that these functions aren't *themselves* handlers, you call them and they *return* a handler. Handler factory functions, if you will.

Here's one that serves up static assets from a directory.

```{r eval=FALSE}
staticHandler <- function(root) {
  force(root)
  return(function(req) {
    if (!identical(req$REQUEST_METHOD, 'GET'))
      return(NULL)

    path <- req$PATH_INFO

    if (is.null(path))
      return(httpResponse(400, content="<h1>Bad Request</h1>"))

    if (path == '/')
      path <- '/index.html'

    abs.path <- resolve(root, path)
    if (is.null(abs.path))
      return(NULL)

    ext <- tools::file_ext(abs.path)
    content.type <- getContentType(ext)
    response.content <- readBin(abs.path, 'raw', n=file.info(abs.path)$size)
    return(httpResponse(200, content.type, response.content))
  })
}
```

## Handler manager

The handler manager gives you a place to register handlers (of both http and websocket varieties) and provides an httpuv-compatible set of callbacks for invoking them.

```{r eval=FALSE}
HandlerManager <- setRefClass("HandlerManager",
  fields = list(
    handlers = "list",
    wsHandlers = "list"
  ),
  methods = list(
    addHandler = function(handler) {
      if (length(handlers) == 0)
        handlers <<- list(handler)
      else
        handlers <<- c(handlers, list(handler))
    },
    addWSHandler = function(wsHandler) {
      if (length(wsHandlers) == 0)
        wsHandlers <<- list(wsHandler)
      else
        wsHandlers <<- c(wsHandlers, list(wsHandler))
    },
    clear = function() {
      handlers <<- list()
      wsHandlers <<- list()
    },
    createHttpuvApp = function() {
      list(
        onHeaders = function(req) {
          maxSize <- getOption('shiny.maxRequestSize', 5 * 1024 * 1024)
          if (maxSize <= 0)
            return(NULL)
          
          reqSize <- 0
          if (length(req$CONTENT_LENGTH) > 0)
            reqSize <- as.numeric(req$CONTENT_LENGTH)
          else if (length(req$HTTP_TRANSFER_ENCODING) > 0)
            reqSize <- Inf
          
          if (reqSize > maxSize) {
            return(list(status = 413L,
              headers = list(
                'Content-Type' = 'text/plain'
              ),
              body = 'Maximum upload size exceeded'))
          }
          else {
            return(NULL)
          }
        },
        call = .httpServer(
          function (req) {
            for (handler in handlers) {
              result <- handler(req)
              if (!is.null(result))
                return(result)
            }
            return(NULL)
          },
          getOption('shiny.sharedSecret', NULL)
        ),
        onWSOpen = function(ws) {
          for (wsHandler in wsHandlers) {
            result <- wsHandler(ws)
            if (!is.null(result))
              return(result)
          }
          return(NULL)
        }
      )
    },
    .httpServer = function(handler, sharedSecret) {
      filter <- getOption('shiny.http.response.filter', NULL)
      if (is.null(filter))
        filter <- function(req, response) response
      
      function(req) {
        if (!is.null(sharedSecret)
          && !identical(sharedSecret, req$HTTP_SHINY_SHARED_SECRET)) {
          return(list(status=403,
            body='<h1>403 Forbidden</h1><p>Shared secret mismatch</p>',
            headers=list('Content-Type' = 'text/html')))
        }
        
        response <- handler(req)
        if (is.null(response))
          response <- httpResponse(404, content="<h1>Not Found</h1>")
        
        headers <- as.list(response$headers)
        headers$'Content-Type' <- response$content_type
        
        response <- filter(req, response)
        return(list(status=response$status,
          body=response$content,
          headers=headers))
      }
    }
  )
)
```
